\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in R

### 2014-01-13 papers/jss/article.Rnw
second pass at data frames
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@765 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[article]{jss}
\usepackage{booktabs}

@@

\section{Evaluation: data.frame to Protocol Buffer Serialization}
\label{sec:evaluation}

The \pkg{RHIPE} package \citep{rhipe} also includes a Protocol integration with R.
However, its implementation takes a different approach: any R object is
serialized into a message based on a single catch-all \texttt{proto} schema.
A similar approach was taken by \pkg{RProtoBufUtils} package (which has now been integrated in
\pkg{RProtoBuf}). Unlike \pkg{RHIPE}, however, \pkg{RProtoBufUtils}
depended upon on, and extended, \pkg{RProtoBuf} for underlying message operations.
%DE Shall this go away now that we sucket RPBUtils into RBP?

One key extension which \pkg{RProtoBufUtils} brought to \pkg{RProtoBuf} is the 
\texttt{serialize\_pb} method to convert R objects into serialized
Protocol Buffers in the catch-all schema. The \texttt{can\_serialize\_pb}
method can then be used to determine whether the given R object can safely
be expressed in this way.  To illustrate how this method works, we
attempt to convert all of the built-in datasets from R into this
serialized Protocol Buffer representation.

@@

There are \Sexpr{n} standard data sets included in the base-r \pkg{datasets}
package. These datasets include data frames, matrices, time series, tables lists,
and some more exotic data classes. The \texttt{can\_serialize\_pb} method can be 
used to determine which of those can fully be converted to the \textt{rexp.proto}
Protocol Buffer representation:

@@

>>>>

\documentclass[article]{jss}
\usepackage{booktabs}
\usepackage[toc,page]{appendix}

@@

\section{Converting R Data Structures into Protocol Buffers}
\label{sec:evaluation}

The previous sections discussed functionality in the \pkg{RProtoBuf} package
for creating, manipulating, parsing and serializing Protocol Buffer messages.
In addition to these low-level methods, the package also has some high level
functionality for automatically converting R data structures into protocol
buffers and vice versa. The \texttt{serialize\_pb} and \texttt{unserialize\_pb}
functions convert arbitrary R objects into a universal Protocol Buffer structure:

<<>>=
msg <- serialize_pb(iris, NULL)
identical(iris, unserialize_pb(msg))
@

In order to accomplish this, \pkg{RProtoBuf} uses the same catch-all \texttt{proto}
schema that \pkg{RHIPE} uses for exchanging R data with Hadoop \citep{rhipe}. This 
schema, which we will refer to as \texttt{rexp.proto} is printed in appendix 
\ref{rex.proto}. Even though the \texttt{RHIPE} implementation is written in Java and
\texttt{RProtoBuf} is writting in R and \texttt{C++}, the Protocol Buffer messages
are naturally compatible between the two systems because they use the same schema. 
This shows the power of using a schema based cross-platform format such as Protocol 
Buffers: interoperability is archieved without tight coordination or collaboration.

\subsection{Evaluation: Converting R Data Sets}

To illustrate how this method works, we attempt to convert all of the built-in 
datasets from R into this serialized Protocol Buffer representation.

@@

There are \Sexpr{n} standard data sets included in the base-r \pkg{datasets}
package. These datasets include data frames, matrices, time series, tables lists,
and some more exotic data classes. The \texttt{can\_serialize\_pb} method can be 
used to determine which of those can fully be converted to the \texttt{rexp.proto}
Protocol Buffer representation:

@@

\newpage
\begin{appendices}

\section{The rexp.proto schema descriptor}
\label{rexp.proto}

Below a print of the \texttt{rexp.proto} schema (originally designed by \cite{rhipe})
that is included with the \pkg{RProtoBuf} package and used by \texttt{serialize\_pb} and
\texttt{unserialize\_pb}.

\begin{verbatim}
package rexp;

message REXP {
  enum RClass {
    STRING = 0;
    RAW = 1;
    REAL = 2;
    COMPLEX = 3;
    INTEGER = 4;
    LIST = 5;
    LOGICAL = 6;
    NULLTYPE = 7;
  }
  enum RBOOLEAN {
    F=0;
    T=1;
    NA=2;
  }

  required RClass rclass = 1 ; 
  repeated double realValue = 2 [packed=true];
  repeated sint32 intValue = 3 [packed=true];
  repeated RBOOLEAN booleanValue = 4;
  repeated STRING stringValue = 5;
  optional bytes rawValue = 6;
  repeated CMPLX complexValue = 7;
  repeated REXP rexpValue = 8;
  repeated string attrName = 11;
  repeated REXP attrValue = 12;
}
message STRING {
  optional string strval = 1;
  optional bool isNA = 2 [default=false];
}
message CMPLX {
  optional double real = 1 [default=0];
  required double imag = 2;
}
\end{verbatim}
\end{appendices}
